Explaining Black-Box Models through Counterfactuals

### 2023-11-17 examples/CounterfactualExplanations.jl/paper.tex
bst seems to have no effect

% Keywords command
\providecommand{\JCONkeywords}[1]
{
  \small	
  \section*{Keywords} #1
}
\usepackage{orcidlink}
\definecolor{mypink}{RGB}{219, 48, 122}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Explaining Black-Box Models through Counterfactuals},
  pdfkeywords={Julia, Explainable Artificial
Intelligence, Counterfactual Explanations, Algorithmic Recourse},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

Machine Learning models like Deep Neural Networks have become so complex
and opaque over recent years that they are generally considered
black-box systems. This lack of transparency exacerbates several other
problems typically associated with these models: they tend to be
unstable (Goodfellow, Shlens, and Szegedy 2014), encode existing biases
(Buolamwini and Gebru 2018) and learn representations that are
surprising or even counter-intuitive from a human perspective
(Buolamwini and Gebru 2018). Nonetheless, they often form the basis for
data-driven decision-making systems in real-world applications.

As others have pointed out, this scenario gives rise to an undesirable
principal-agent problem involving a group of principals---i.e.~human
stakeholders---that fail to understand the behaviour of their
agent---i.e.~the black-box system (Borch 2022). The group of principals
may include programmers, product managers and other decision-makers who
develop and operate the system as well as those individuals ultimately
subject to the decisions made by the system. In practice, decisions made
by black-box systems are typically left unchallenged since the group of
principals cannot scrutinize them:

\begin{quote}
``You cannot appeal to (algorithms). They do not listen. Nor do they
bend.'' (O'Neil 2016)
\end{quote}

@@

The field of XAI is still relatively young and made up of a variety of
subdomains, definitions, concepts and taxonomies. Covering all of these
is beyond the scope of this article, so we will focus only on high-level
concepts. The following literature surveys provide more detail: Arrieta
et al.~(2020) provide a broad overview of XAI (Arrieta et al. 2020); Fan
et al.~(2020) focus on explainability in the context of deep learning
(Fan, Xiong, and Wang 2020); and finally, Karimi et al.~(2020) (Karimi,
Barthe, et al. 2020) and Verma et al.~(2020) Verma, Dickerson, and Hines
(2020) offer detailed reviews of the literature on Counterfactual
Explanations and Algorithmic Recourse (see also Molnar (2020) and
Varshney (2022)). Miller (2019) explicitly discusses the concept of
explainability from the perspective of a social scientist (Miller 2019).

The first broad distinction we want to make here is between
\textbf{Interpretable} and \textbf{Explainable} AI. These terms are
often used interchangeably, but this can lead to confusion. We find the
distinction made in Rudin (2019) useful: Interpretable AI involves
models that are inherently interpretable and transparent such as general
additive models (GAM), decision trees and rule-based models; Explainable
AI involves models that are not inherently interpretable but require
additional tools to be explainable to humans. Examples of the latter
include Ensembles, Support Vector Machines and Deep Neural Networks.
Some would argue that we best avoid the second category of models
altogether and instead focus solely on interpretable AI Rudin (2019).
While we agree that initial efforts should always be geared towards
interpretable models, avoiding black boxes altogether would entail
missed opportunities and anyway is probably not very realistic at this
point. For that reason, we expect the need for XAI to persist in the
medium term. Explainable AI can further be broadly divided into
\textbf{global} and \textbf{local} explainability: the former is
concerned with explaining the average behaviour of a model, while the
latter involves explanations for individual predictions (Molnar 2020).
Tools for global explainability include partial dependence plots (PDP),
which involve the computation of marginal effects through Monte Carlo,
and global surrogates. A surrogate model is an interpretable model that
is trained to explain the predictions of a black-box model.

@@

A recent study has shown that even experienced data scientists tend to
put too much trust in explanations produced by LIME and SHAP (Kaur et
al. 2020). Another recent work has shown that both methods can be easily
fooled: they depend on random input perturbations, a property that can
be abused by adverse agents to essentially whitewash strongly biased
black-box models (Slack et al. 2020). In related work, the same authors
find that while gradient-based Counterfactual Explanations can also be
manipulated, there is a straightforward way to protect against this in
practice (Slack et al. 2021). In the context of quality assessment, it
is also worth noting that---contrary to surrogate explainers---CE always
achieve full fidelity by construction: counterfactuals are searched with
respect to the black-box classifier, not some proxy for it. That being
said, CE should also be used with care and research around them is still
in its early stages.

@@

Counterfactual search involves feature perturbations: we are interested
in understanding how we need to change individual attributes in order to
change the model output to a desired value or label (Molnar 2020).
Typically the underlying methodology is presented in the context of
binary classification: \(M: \mathcal{X} \mapsto \mathcal{Y}\) where
\(\mathcal{X}\subset\mathbb{R}^D\) and \(\mathcal{Y}=\{0,1\}\). Further,
let \(t=1\) be the target class and let \(x\) denote the factual feature
vector of some individual sample outside of the target class, so
\(y=M(x)=0\). We follow this convention here, though it should be noted
that the ideas presented here also carry over to multi-class problems
and regression (Molnar 2020).

The counterfactual search objective originally proposed by Wachter,
Mittelstadt, and Russell (2017) is as follows

@@

Numerous extensions to this simple approach have been developed since CE
were first proposed in 2017 (see Verma, Dickerson, and Hines (2020) and
(Karimi, Barthe, et al. 2020) for surveys). The various approaches
largely differ in that they use different flavours of search objective
defined in Equation~\ref{eq-solution}. Different penalties are often
used to address many of the desirable properties of effective CE that
have been set out. These desiderata include: \textbf{proximity} --- the
distance between factual and counterfactual features should be small
(Wachter, Mittelstadt, and Russell 2017); \textbf{actionability} --- the
proposed recourse should be actionable Poyiadzi et al. (2020);
\textbf{plausibility} --- the counterfactual explanation should be
plausible to a human Schut et al. (2021); \textbf{sparsity} --- the
counterfactual explanation should involve as few individual feature
changes as possible (Schut et al. 2021); \textbf{robustness} --- the
counterfactual explanation should be robust to domain and model shifts
(Upadhyay, Joshi, and Lakkaraju 2021); \textbf{diversity} --- ideally
multiple diverse counterfactuals should be provided (Mothilal, Sharma,
and Tan 2020); and \textbf{causality} --- counterfactuals should respect
the structural causal model underlying the data generating process
Karimi, Schölkopf, and Valera (2021).

@@

To the best of our knowledge, the package introduced here provides the
first implementation of Counterfactual Explanations in Julia and
therefore represents a novel contribution to the community. As for other
programming languages, we are only aware of one other unifying
framework: the Python library
\href{https://carla-counterfactual-and-recourse-library.readthedocs.io/en/latest/?badge=latest}{CARLA}
(Pawelczyk et al. 2021).\footnote{While we were writing this paper, the
  \texttt{R} package \texttt{counterfactuals} was released (Dandl et al.
  2023). The developers seem to also envision a unifying framework, but
  the project appears to still be in its early stages.} In addition to
that, there exists open-source code for some specific approaches to CE
that have been proposed in recent years. The approach-specific
implementations that we have been able to find are generally
well-documented, but exclusively in Python. For example, a PyTorch
implementation of a greedy generator for Bayesian models proposed in
Schut et al. (2021) has been released. As another example, the popular
\href{https://github.com/interpretml}{InterpretML} library includes an
implementation of a diverse counterfactual generator (Mothilal, Sharma,
and Tan 2020).

@@

The package currently offers native support for models built and trained
in \href{https://fluxml.ai/}{Flux} (Innes 2018) as well as a small
subset of models made available through
\href{https://alan-turing-institute.github.io/MLJ.jl/dev/}{MLJ} (Blaom
et al. 2020). While in general it is assumed that users resort to this
package to explain their pre-trained models, we provide a simple API
call to train the following
\href{https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/tutorials/model_catalogue/}{models}:

\begin{itemize}
\tightlist
\item
  Linear Classifier (Logistic Regression and Multinomial Logit)
\item
  Multi-Layer Perceptron (Deep Neural Network)
\item
  Deep Ensemble Lakshminarayanan, Pritzel, and Blundell (2016)
\item
  Decision Tree, Random Forest, Gradient Boosted Trees
\end{itemize}

@@

A large and growing number of counterfactual
\href{https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/explanation/generators/overview/}{generators}
have already been implemented in our package (Table~\ref{tbl-gen}). At a
high level, we distinguish generators in terms of their compatible model
types, their default search space, and their composability. All
``gradient-based'' generators are compatible with differentiable models,
e.g.~\texttt{Flux} and \texttt{torch}, while ``tree-based'' generators
are only applicable to models that involve decision trees. Concerning
the search space, it is possible to search counterfactuals in a
lower-dimensional latent embedding of the feature space that implicitly
encodes the data-generating process (DGP). To learn the latent
embedding, existing work has typically relied on generative models or
existing causal knowledge Karimi, Schölkopf, and Valera (2021). While
this notion is compatible with all of our gradient-based generators,
only some generators search a latent space by default. Finally,
composability implies that the given generator can be blended with any
other composable generator, which we discuss in
Section~\ref{sec-gen-comp}.

Beyond these broad technical distinctions, generators largely differ in
terms of how they address the various desiderata mentioned above:
\emph{ClapROAR} aims to preserve the classifier, i.e.~to generate
counterfactuals that are robust to endogenous model shifts (Altmeyer et
al. 2023); \emph{CLUE} searches plausible counterfactuals in the latent
embedding of a generative model by explicitly minimising predictive
entropy (Antorán et al. 2020); \emph{DiCE} is designed to generate
multiple, maximally diverse counterfactuals (Mothilal, Sharma, and Tan
2020); \emph{FeatureTweak} leverages the internals of decision trees to
search counterfactuals on a feature-by-feature basis, finding the
counterfactual that tweaks the features in the least costly way (Tolomei
et al. 2017); \emph{Gravitational} aims to generate plausible and robust
counterfactuals by minimising the distance to observed samples in the
target class (Altmeyer et al. 2023); \emph{Greedy} aims to generate
plausible counterfactuals by implicitly minimising predictive
uncertainty of Bayesian classifiers (Schut et al. 2021);
\emph{GrowingSpheres} is model-agnostic, relying solely on identifying
nearest neighbours of counterfactuals in the target class by gradually
increasing the search radius and then moving counterfactuals in that
direction(Laugel et al. 2017); \emph{PROBE} generates probabilistically
robust counterfactuals (Pawelczyk et al. 2022); \emph{REVISE} addresses
the need for plausibility by searching counterfactuals in the latent
embedding of a Variational Autoencoder (VAE) (Joshi et al. 2019);
\emph{Wachter} is the baseline approach that only penalises the distance
to the original sample (Wachter, Mittelstadt, and Russell 2017).

\onecolumn

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2500}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2500}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2500}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2500}}@{}}
\caption{Overview of implemented counterfactual
generators.}\label{tbl-gen}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Generator
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Model Type
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Search Space
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Composable
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Generator
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Model Type
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Search Space
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Composable
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
ClaPROAR (Altmeyer et al. 2023) & gradient based & feature & yes \\
CLUE (Antorán et al. 2020) & gradient based & latent & yes \\
DiCE (Mothilal, Sharma, and Tan 2020) & gradient based & feature &
yes \\
FeatureTweak (Tolomei et al. 2017) & tree based & feature & no \\
Gravitational (Altmeyer et al. 2023) & gradient based & feature & yes \\
Greedy (Schut et al. 2021) & gradient based & feature & yes \\
GrowingSpheres (Laugel et al. 2017) & agnostic & feature & no \\
PROBE (Pawelczyk et al. 2022) & gradient based & feature & no \\
REVISE (Joshi et al. 2019) & gradient based & latent & yes \\
Wachter (Wachter, Mittelstadt, and Russell 2017) & gradient based &
feature & yes \\
\end{longtable}

\twocolumn

@@

\begin{itemize}
\tightlist
\item
  Adult Census (Barry Becker 1996)
\item
  California Housing (Pace and Barry 1997)
\item
  CIFAR10 (Krizhevsky 2009)
\item
  German Credit (Hoffman 1994)
\item
  Give Me Some Credit (Kaggle 2011)
\item
  MNIST (LeCun 1998) and Fashion MNIST (Xiao, Rasul, and Vollgraf 2017)
\item
  UCI defaultCredit (Yeh and Lien 2009)
\end{itemize}

@@

The package also makes it easy to
\href{https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/tutorials/evaluation/}{evaluate}
counterfactuals with respect to many of the desiderata mentioned above.
For example, users may want to infer how costly the provided recourse is
to individuals. To this end, we can measure the distance of the
counterfactual from its original value. The API call to compute the
distance metric defined in Wachter, Mittelstadt, and Russell (2017), for
instance, is as simple as \texttt{evaluate(ce;\ measure=distance\_mad)},
where \texttt{ce} can also be a vector of
\texttt{CounterfactualExplanation}s.

@@

\end{document}

>>>>

% Keywords command
\providecommand{\JCONkeywords}[1]
{
  \small	
  \section*{Keywords} #1
}

% Two-column table
% Lifted from https://github.com/quarto-journals/elsevier/blob/main/_extensions/elsevier/partials/_two-column-longtable.tex.
\usepackage{float}
\makeatletter
\let\oldlt\longtable
\let\endoldlt\endlongtable
\def\longtable{\@ifnextchar[\longtable@i \longtable@ii}
\def\longtable@i[#1]{\begin{figure}[H]
\onecolumn
\begin{minipage}{0.5\textwidth}
\oldlt[#1]
}
\def\longtable@ii{\begin{figure}[H]
\onecolumn
\begin{minipage}{0.5\textwidth}
\oldlt
}
\def\endlongtable{\endoldlt
\end{minipage}
\twocolumn
\end{figure}}
\makeatother

% Remove whitespace after paragraphs
\setlength{\parskip}{0pt}

\usepackage{orcidlink}
\definecolor{mypink}{RGB}{219, 48, 122}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Explaining Black-Box Models through Counterfactuals},
  pdfkeywords={Julia, Explainable Artificial
Intelligence, Counterfactual Explanations, Algorithmic Recourse},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

Machine Learning models like Deep Neural Networks have become so complex
and opaque over recent years that they are generally considered
black-box systems. This lack of transparency exacerbates several other
problems typically associated with these models: they tend to be
unstable (\citeproc{ref-goodfellow2014explaining}{Goodfellow, Shlens,
and Szegedy 2014}), encode existing biases
(\citeproc{ref-buolamwini2018gender}{Buolamwini and Gebru 2018}) and
learn representations that are surprising or even counter-intuitive from
a human perspective (\citeproc{ref-buolamwini2018gender}{Buolamwini and
Gebru 2018}). Nonetheless, they often form the basis for data-driven
decision-making systems in real-world applications.

As others have pointed out, this scenario gives rise to an undesirable
principal-agent problem involving a group of principals---i.e.~human
stakeholders---that fail to understand the behaviour of their
agent---i.e.~the black-box system (\citeproc{ref-borch2022machine}{Borch
2022}). The group of principals may include programmers, product
managers and other decision-makers who develop and operate the system as
well as those individuals ultimately subject to the decisions made by
the system. In practice, decisions made by black-box systems are
typically left unchallenged since the group of principals cannot
scrutinize them:

\begin{quote}
``You cannot appeal to (algorithms). They do not listen. Nor do they
bend.'' (\citeproc{ref-oneil2016weapons}{O'Neil 2016})
\end{quote}

@@

The field of XAI is still relatively young and made up of a variety of
subdomains, definitions, concepts and taxonomies. Covering all of these
is beyond the scope of this article, so we will focus only on high-level
concepts. The following literature surveys provide more detail: Arrieta
et al.~(2020) provide a broad overview of XAI
(\citeproc{ref-arrieta2020explainable}{Arrieta et al. 2020}); Fan et
al.~(2020) focus on explainability in the context of deep learning
(\citeproc{ref-fan2020interpretability}{Fan, Xiong, and Wang 2020}); and
finally, Karimi et al.~(2020) (\citeproc{ref-karimi2020survey}{Karimi,
Barthe, et al. 2020}) and Verma et al.~(2020) Verma, Dickerson, and
Hines (\citeproc{ref-verma2020counterfactual}{2020}) offer detailed
reviews of the literature on Counterfactual Explanations and Algorithmic
Recourse (see also Molnar (\citeproc{ref-molnar2020interpretable}{2020})
and Varshney (\citeproc{ref-varshney2022trustworthy}{2022})). Miller
(2019) explicitly discusses the concept of explainability from the
perspective of a social scientist
(\citeproc{ref-miller2019explanation}{Miller 2019}).

The first broad distinction we want to make here is between
\textbf{Interpretable} and \textbf{Explainable} AI. These terms are
often used interchangeably, but this can lead to confusion. We find the
distinction made in Rudin (\citeproc{ref-rudin2019stop}{2019}) useful:
Interpretable AI involves models that are inherently interpretable and
transparent such as general additive models (GAM), decision trees and
rule-based models; Explainable AI involves models that are not
inherently interpretable but require additional tools to be explainable
to humans. Examples of the latter include Ensembles, Support Vector
Machines and Deep Neural Networks. Some would argue that we best avoid
the second category of models altogether and instead focus solely on
interpretable AI Rudin (\citeproc{ref-rudin2019stop}{2019}). While we
agree that initial efforts should always be geared towards interpretable
models, avoiding black boxes altogether would entail missed
opportunities and anyway is probably not very realistic at this point.
For that reason, we expect the need for XAI to persist in the medium
term. Explainable AI can further be broadly divided into \textbf{global}
and \textbf{local} explainability: the former is concerned with
explaining the average behaviour of a model, while the latter involves
explanations for individual predictions
(\citeproc{ref-molnar2020interpretable}{Molnar 2020}). Tools for global
explainability include partial dependence plots (PDP), which involve the
computation of marginal effects through Monte Carlo, and global
surrogates. A surrogate model is an interpretable model that is trained
to explain the predictions of a black-box model.

@@

A recent study has shown that even experienced data scientists tend to
put too much trust in explanations produced by LIME and SHAP
(\citeproc{ref-kaur2020interpreting}{Kaur et al. 2020}). Another recent
work has shown that both methods can be easily fooled: they depend on
random input perturbations, a property that can be abused by adverse
agents to essentially whitewash strongly biased black-box models
(\citeproc{ref-slack2020fooling}{Slack et al. 2020}). In related work,
the same authors find that while gradient-based Counterfactual
Explanations can also be manipulated, there is a straightforward way to
protect against this in practice
(\citeproc{ref-slack2021counterfactual}{Slack et al. 2021}). In the
context of quality assessment, it is also worth noting that---contrary
to surrogate explainers---CE always achieve full fidelity by
construction: counterfactuals are searched with respect to the black-box
classifier, not some proxy for it. That being said, CE should also be
used with care and research around them is still in its early stages.

@@

Counterfactual search involves feature perturbations: we are interested
in understanding how we need to change individual attributes in order to
change the model output to a desired value or label
(\citeproc{ref-molnar2020interpretable}{Molnar 2020}). Typically the
underlying methodology is presented in the context of binary
classification: \(M: \mathcal{X} \mapsto \mathcal{Y}\) where
\(\mathcal{X}\subset\mathbb{R}^D\) and \(\mathcal{Y}=\{0,1\}\). Further,
let \(t=1\) be the target class and let \(x\) denote the factual feature
vector of some individual sample outside of the target class, so
\(y=M(x)=0\). We follow this convention here, though it should be noted
that the ideas presented here also carry over to multi-class problems
and regression (\citeproc{ref-molnar2020interpretable}{Molnar 2020}).

The counterfactual search objective originally proposed by Wachter,
Mittelstadt, and Russell
(\citeproc{ref-wachter2017counterfactual}{2017}) is as follows

@@

Numerous extensions to this simple approach have been developed since CE
were first proposed in 2017 (see Verma, Dickerson, and Hines
(\citeproc{ref-verma2020counterfactual}{2020}) and
(\citeproc{ref-karimi2020survey}{Karimi, Barthe, et al. 2020}) for
surveys). The various approaches largely differ in that they use
different flavours of search objective defined in
Equation~\ref{eq-solution}. Different penalties are often used to
address many of the desirable properties of effective CE that have been
set out. These desiderata include: \textbf{proximity} --- the distance
between factual and counterfactual features should be small
(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
Russell 2017}); \textbf{actionability} --- the proposed recourse should
be actionable Poyiadzi et al. (\citeproc{ref-poyiadzi2020face}{2020});
\textbf{plausibility} --- the counterfactual explanation should be
plausible to a human Schut et al.
(\citeproc{ref-schut2021generating}{2021}); \textbf{sparsity} --- the
counterfactual explanation should involve as few individual feature
changes as possible (\citeproc{ref-schut2021generating}{Schut et al.
2021}); \textbf{robustness} --- the counterfactual explanation should be
robust to domain and model shifts
(\citeproc{ref-upadhyay2021robust}{Upadhyay, Joshi, and Lakkaraju
2021}); \textbf{diversity} --- ideally multiple diverse counterfactuals
should be provided (\citeproc{ref-mothilal2020explaining}{Mothilal,
Sharma, and Tan 2020}); and \textbf{causality} --- counterfactuals
should respect the structural causal model underlying the data
generating process Karimi, Schölkopf, and Valera
(\citeproc{ref-karimi2021algorithmic}{2021}).

@@

To the best of our knowledge, the package introduced here provides the
first implementation of Counterfactual Explanations in Julia and
therefore represents a novel contribution to the community. As for other
programming languages, we are only aware of one other unifying
framework: the Python library
\href{https://carla-counterfactual-and-recourse-library.readthedocs.io/en/latest/?badge=latest}{CARLA}
(\citeproc{ref-pawelczyk2021carla}{Pawelczyk et al. 2021}).\footnote{While
  we were writing this paper, the \texttt{R} package
  \texttt{counterfactuals} was released
  (\citeproc{ref-dandl2023counterfactuals}{Dandl et al. 2023}). The
  developers seem to also envision a unifying framework, but the project
  appears to still be in its early stages.} In addition to that, there
exists open-source code for some specific approaches to CE that have
been proposed in recent years. The approach-specific implementations
that we have been able to find are generally well-documented, but
exclusively in Python. For example, a PyTorch implementation of a greedy
generator for Bayesian models proposed in Schut et al.
(\citeproc{ref-schut2021generating}{2021}) has been released. As another
example, the popular \href{https://github.com/interpretml}{InterpretML}
library includes an implementation of a diverse counterfactual generator
(\citeproc{ref-mothilal2020explaining}{Mothilal, Sharma, and Tan 2020}).

@@

The package currently offers native support for models built and trained
in \href{https://fluxml.ai/}{Flux} (\citeproc{ref-innes2018flux}{Innes
2018}) as well as a small subset of models made available through
\href{https://alan-turing-institute.github.io/MLJ.jl/dev/}{MLJ}
(\citeproc{ref-blaom2020mlj}{Blaom et al. 2020}). While in general it is
assumed that users resort to this package to explain their pre-trained
models, we provide a simple API call to train the following
\href{https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/tutorials/model_catalogue/}{models}:

\begin{itemize}
\tightlist
\item
  Linear Classifier (Logistic Regression and Multinomial Logit)
\item
  Multi-Layer Perceptron (Deep Neural Network)
\item
  Deep Ensemble Lakshminarayanan, Pritzel, and Blundell
  (\citeproc{ref-lakshminarayanan2016simple}{2016})
\item
  Decision Tree, Random Forest, Gradient Boosted Trees
\end{itemize}

@@

A large and growing number of counterfactual
\href{https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/explanation/generators/overview/}{generators}
have already been implemented in our package (Table~\ref{tbl-gen}). At a
high level, we distinguish generators in terms of their compatible model
types, their default search space, and their composability. All
``gradient-based'' generators are compatible with differentiable models,
e.g.~\texttt{Flux} and \texttt{torch}, while ``tree-based'' generators
are only applicable to models that involve decision trees. Concerning
the search space, it is possible to search counterfactuals in a
lower-dimensional latent embedding of the feature space that implicitly
encodes the data-generating process (DGP). To learn the latent
embedding, existing work has typically relied on generative models or
existing causal knowledge Karimi, Schölkopf, and Valera
(\citeproc{ref-karimi2021algorithmic}{2021}). While this notion is
compatible with all of our gradient-based generators, only some
generators search a latent space by default. Finally, composability
implies that the given generator can be blended with any other
composable generator, which we discuss in Section~\ref{sec-gen-comp}.

Beyond these broad technical distinctions, generators largely differ in
terms of how they address the various desiderata mentioned above:
\emph{ClapROAR} aims to preserve the classifier, i.e.~to generate
counterfactuals that are robust to endogenous model shifts
(\citeproc{ref-altmeyer2023endogenous}{Altmeyer et al. 2023});
\emph{CLUE} searches plausible counterfactuals in the latent embedding
of a generative model by explicitly minimising predictive entropy
(\citeproc{ref-antoran2020getting}{Antorán et al. 2020}); \emph{DiCE} is
designed to generate multiple, maximally diverse counterfactuals
(\citeproc{ref-mothilal2020explaining}{Mothilal, Sharma, and Tan 2020});
\emph{FeatureTweak} leverages the internals of decision trees to search
counterfactuals on a feature-by-feature basis, finding the
counterfactual that tweaks the features in the least costly way
(\citeproc{ref-tolomei2017interpretable}{Tolomei et al. 2017});
\emph{Gravitational} aims to generate plausible and robust
counterfactuals by minimising the distance to observed samples in the
target class (\citeproc{ref-altmeyer2023endogenous}{Altmeyer et al.
2023}); \emph{Greedy} aims to generate plausible counterfactuals by
implicitly minimising predictive uncertainty of Bayesian classifiers
(\citeproc{ref-schut2021generating}{Schut et al. 2021});
\emph{GrowingSpheres} is model-agnostic, relying solely on identifying
nearest neighbours of counterfactuals in the target class by gradually
increasing the search radius and then moving counterfactuals in that
direction(\citeproc{ref-laugel2017inversea}{Laugel et al. 2017});
\emph{PROBE} generates probabilistically robust counterfactuals
(\citeproc{ref-pawelczyk2022probabilistically}{Pawelczyk et al. 2022});
\emph{REVISE} addresses the need for plausibility by searching
counterfactuals in the latent embedding of a Variational Autoencoder
(VAE) (\citeproc{ref-joshi2019realistic}{Joshi et al. 2019});
\emph{Wachter} is the baseline approach that only penalises the distance
to the original sample
(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
Russell 2017}).

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2500}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2500}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2500}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2500}}@{}}
\caption{Overview of implemented counterfactual
generators.}\label{tbl-gen}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Generator
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Model Type
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Search Space
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Composable
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Generator
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Model Type
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Search Space
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Composable
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
ClaPROAR (\citeproc{ref-altmeyer2023endogenous}{Altmeyer et al. 2023}) &
gradient based & feature & yes \\
CLUE (\citeproc{ref-antoran2020getting}{Antorán et al. 2020}) & gradient
based & latent & yes \\
DiCE (\citeproc{ref-mothilal2020explaining}{Mothilal, Sharma, and Tan
2020}) & gradient based & feature & yes \\
FeatureTweak (\citeproc{ref-tolomei2017interpretable}{Tolomei et al.
2017}) & tree based & feature & no \\
Gravitational (\citeproc{ref-altmeyer2023endogenous}{Altmeyer et al.
2023}) & gradient based & feature & yes \\
Greedy (\citeproc{ref-schut2021generating}{Schut et al. 2021}) &
gradient based & feature & yes \\
GrowingSpheres (\citeproc{ref-laugel2017inversea}{Laugel et al. 2017}) &
agnostic & feature & no \\
PROBE (\citeproc{ref-pawelczyk2022probabilistically}{Pawelczyk et al.
2022}) & gradient based & feature & no \\
REVISE (\citeproc{ref-joshi2019realistic}{Joshi et al. 2019}) & gradient
based & latent & yes \\
Wachter (\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt,
and Russell 2017}) & gradient based & feature & yes \\
\end{longtable}

@@

\begin{itemize}
\tightlist
\item
  Adult Census (\citeproc{ref-becker1996adult}{Barry Becker 1996})
\item
  California Housing (\citeproc{ref-pace1997sparse}{Pace and Barry
  1997})
\item
  CIFAR10 (\citeproc{ref-krizhevsky2009learning}{Krizhevsky 2009})
\item
  German Credit (\citeproc{ref-hoffman1994german}{Hoffman 1994})
\item
  Give Me Some Credit (\citeproc{ref-kaggle2011give}{Kaggle 2011})
\item
  MNIST (\citeproc{ref-lecun1998mnist}{LeCun 1998}) and Fashion MNIST
  (\citeproc{ref-xiao2017fashion}{Xiao, Rasul, and Vollgraf 2017})
\item
  UCI defaultCredit (\citeproc{ref-yeh2009comparisons}{Yeh and Lien
  2009})
\end{itemize}

@@

The package also makes it easy to
\href{https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/tutorials/evaluation/}{evaluate}
counterfactuals with respect to many of the desiderata mentioned above.
For example, users may want to infer how costly the provided recourse is
to individuals. To this end, we can measure the distance of the
counterfactual from its original value. The API call to compute the
distance metric defined in Wachter, Mittelstadt, and Russell
(\citeproc{ref-wachter2017counterfactual}{2017}), for instance, is as
simple as \texttt{evaluate(ce;\ measure=distance\_mad)}, where
\texttt{ce} can also be a vector of \texttt{CounterfactualExplanation}s.

@@

% **************GENERATED FILE, DO NOT EDIT**************
